Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-01-09 analysis/paper.qmd
Correct and refactor archaeo_pred

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 81<!-- TODO update --> plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  Modelled ranges do not reliably predict the observed  occurrence of specific 
  species at archaeological sites,
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Stop modelling
- [ ] First draft:
  - [x] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [x] Discussion
  - [ ] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
- [x] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

@@

The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@cite].
Even before the end of the Ice Age, these species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@cite] and eventually pre-domestication cultivation [@cite].
As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.

@@

Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
In the first sense, it has been used most extensively to model the range of humans and other hominin species [e.g. @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
These applications can be distinguihed from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
In an early application to West Asia, @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik? --> [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [@CremaEtAl].

@@

The alternative approach is to train the model using contemporary occurrence and environmental data and then use palaeoenvironmental data to 'hindcast' its predictions backwards in time.
Like @FranklinEtAl2015, we view the hindcasting approach as more promising, because training datasets for both occurrences and environment are far more readily available, complete and reliable for the present than the past.
There is some scepticism in the ecological niche modelling literature about the ability of such models to make accurate predictions in unknown environments (like the past) [@CITES?], 
but here the hindcasting approach also presents an opportunity:
it reserves archaeological occurrence data as an independent dataset that can be used to assess the retrodictive performance of the model.
This possibily was suggested by @FranklinEtAl2015 but to our knowledge our study represents the first attempt to actually do so.

@@

```{r data-archaeo-flora}
# Flora present at archaeological sites
archaeo_flora <- read_tsv(
  raw_data("swasia_neolithic_flora.tsv"),
  col_types = cols(
    source = col_character(),
    source_id = col_character(),
    source_site_name = col_character(),
    site_name = col_character(),
    latitude = col_double(),
    longitude = col_double(),
    phase = col_character(),
    phase_description = col_character(),
    phase_code = col_character(),
    age_start = col_integer(),
    age_end = col_integer(),
    taxon_source = col_character(),
    n = col_double(),
    prop = col_double(),
    reference = col_character(),
    taxon_detail = col_character(),
    taxon = col_character(),
    genus = col_character(),
    family = col_character(),
    category = col_character(),
    founder_crop = col_character(),
    edibility = col_character(),
    grass_type = col_character(),
    legume_type = col_character()
  )
) |>
  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
  # Assign to archaeological periods
  mutate(
    age_mid = age_end + ((age_start - age_end) / 2),
    period = cut(
      -age_mid, 
      breaks = -c(archaeo_periods$start_bp, 0), 
      labels = archaeo_periods$period,
      ordered_result = TRUE
    ),
    climate_period = cut(
      -age_mid, 
      breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
      labels = climate_periods[climate_periods$code != "cur",]$code,
      ordered_result = TRUE
    )
  ) |>
  # Filter to Epipal/Neo
  filter(period %in% epipal | period %in% ppn)

n_archaeo_assemb <- nrow(distinct(archaeo_flora, site_name, phase_code))
n_archaeo_site <- nrow(distinct(archaeo_flora, site_name))
```

```{r data-archaeo-sites}
archaeo_sites <- summarise(
  archaeo_flora, 
  latitude = first(latitude), 
  longitude = first(longitude), 
  .by = c(site_name, period)
) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = latlong, remove = FALSE)
```

@@

```{r data-flora}
# TODO: nest archaeo flora in flora to make it easier to keep track of taxon transformations
if (!file_exists(derived_data("flora.tsv"))) {
  message("Regenerating flora...")
  # Subset of flora to be modelled
  archaeo_flora |>
    group_by(taxon) |>
    summarise(
      n_present = length(unique(phase_code)),
      p_present = n_present / n_archaeo_assemb,
      .groups = "drop"
    ) |>
    filter(n_present > min_n_archaeo) |>
    # Exclude taxa not identified to species
    filter(!str_ends(taxon, "sp.")) ->
    flora

  # Standardise taxonomy
  flora <- flora |>
    # Normalise species names and match to the GBIF Backbone Taxonomy
    mutate(
      taxon = case_match(taxon,
                         "Vicia narbonense" ~ "Vicia narbonensis", # N=3
                         "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
                         "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
                         "Lens orientalis" ~ "Vicia orientalis", # N=2
                         "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
                         "Stipa capensis" ~ "Stipa dregeana", # N=2
                         .default = taxon),
      taxon_original = taxon,
      taxon = map_chr(taxon, gbif_accepted_name)
    ) |>
    # Recode domesticates as their wild progenitors.
    mutate(
      taxon = case_match(
        taxon,
        "Cicer arietinum" ~ "Cicer reticulatum",
        "Hordeum vulgare" ~ "Hordeum spontaneum",
        "Vicia lens" ~ "Vicia orientalis",
        "Linum usitatissimum" ~ "Linum bienne",
        "Triticum aestivum" ~ "Triticum turgidum dicoccum",
        "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
        "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
        "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
        .default = taxon
      )
    ) |>
    # Add wild progenitors of wheats
    bind_rows(
      tibble(
        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"),
        taxon_original = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
      )
    ) |>
    # Deduplicate by canonical taxon
    summarise(
      taxon_detail = paste0("*", taxon_original[taxon_original != taxon], "*", collapse = ", "),
      taxon_detail = na_if(taxon_detail, "**"),
      n_present = max(n_present),
      p_present = max(p_present),
      .by = taxon
    )

  write_tsv(flora, derived_data("flora.tsv"))
}

flora <- read_tsv(
  derived_data("flora.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double()
  )
)
```

The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
Our starting point was a list of `r nrow(flora)` taxa (@tbl-taxa) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
This was based on dataset of `r n_archaeo_assemb` assemblages collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@COMPAG].
We did not attempt to distinguish between the source of the remains;
archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
However, we assume that there presence at a site of human settlement at least implies that they were part of the wider ecosystem that supported habitation there.

Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-taxa).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.

@@

```{r data-n-occ} 
if (!file_exists(derived_data("flora_n_occ.tsv"))) {
  message("Reacquiring GBIF occurrence counts...")
  flora <- flora |>
    mutate(
      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                            hasCoordinate = TRUE, 
                                            hasGeospatialIssue = FALSE,
                                            geometry = bbox_wkt(w_asia)))
    )
  write_tsv(flora, derived_data("flora_n_occ.tsv"))
}

flora <- read_tsv(
  derived_data("flora_n_occ.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_integer()
  )
)

@@

```{r data-occ}
# Cleaning pipeline based on:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
if (!file_exists(derived_data("flora_occ.tsv"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, filter, occurrenceStatus  == "PRESENT"),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 1000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    }),
    # Update occ count after cleaning
    n_occ = map_int(occ, nrow)
  )

@@

  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
}

flora <- read_tsv(
  derived_data("flora_occ.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_integer(),
    gbif_key = col_double(),
    scientific_name = col_character(),
    genus = col_character(),
    species = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
) |>
  nest(occ = gbif_key:latitude)
```

Georeferenced occurrence data for West Asia was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
GBIF was cleaned to removed fossil occurrences, recorded absences, and records with missing, imprecise (>1 km uncertainty), or invalid coordinates.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora, n_occ > min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what accounted for such a disproportionately high number of records.

@@

## Predictor data

```{r tbl-predictors}
# TODO
```

We modelled the occurrence of species as a function of <!-- TODO -->X spatial predictor variables (@tbl-predictors).
These included:

@@

Predictor data was transformed to common equal-area projection and resolution of `r pred_res / 1000` km.

```{r tbl-climate-periods}
#| fig-cap: Palaeoclimatic periods used for hindcasting, after @BrownEtAl2018
# TODO: integrate into text
climate_periods |>
  mutate(
    code = toupper(code),
    age_start = start_bp / 1000,
    age_end = end_bp / 1000
  ) |>
  select(period, code, age_start, age_end) |>
  gt() |>
  cols_label(period = "Period", age_start = "Age") |>
  cols_merge(
    c(period, code),
    pattern = "{1} ({2})"
  ) |>
  cols_merge_range(age_start, age_end) |>
  cols_units(age_start = "ka") |>
  sub_missing()
```

For hindcasting, we used reconstructed bioclimatic data for `r english(nrow(climate_periods) - 1)` key periods (@tbl-climate-periods) generated from downscaled paleoclimate simulations from the HadCM3 general circulation model [@BrownEtAl2018].
Terrain and soil predictors were held constant, since reconstructions of these variables in the past are not available at sufficient scale.
It is not likely that either macroscale topography or soil characteristics have altered significantly over the period of time considered here, so we assume that this does not degrade model performance, and may in fact benefit it by providing 'anchoring' predictors that are independent of climate change.

@@

```{r niche-data}
# Split training, test (for model assessment) and cross-validation sets (for
# tuning)
flora <- flora |>
  mutate(
    # De-sf
    occ = map(occ, \(x) select(tibble(x), -geometry)),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    # occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
    # occ_folds = map(occ_split, \(x) vfold_cv(training(x), v = 10, strata = present))
  )

@@

For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our X<!--TODO--> predictor variables (@tbl-predictors).
We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (¾) and test (¼) partitions.

@@

We trained Random Forest models for `r nrow(flora)` taxa using contemporary occurrence data from GBIF, a random sample of background points, and the predictor variables described in @tbl-predictors.
Substituting the "current" climate predictors for those derived from palaeoclimatic simulations [@BrownEtAl2018], we could then generate hindcast predictions for reconstructed past environments in `r english(nrow(climate_periods))` key climate periods – a total of `r nrow(flora) * nrow(climate_periods)` modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the appendix and accompanying material.

@@

```{r fig.height=10, fig-palaeodist-barely-levant, eval=FALSE, cache=TRUE}
#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
palaeodist_hordeum <- do.call(c, c(
  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
))

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 82<!-- TODO update --> plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  Modelled ranges do not reliably predict the observed  occurrence of specific 
  species at archaeological sites,
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

<!--

* [x] Stop modelling
- [ ] First draft:
  - [x] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [x] Discussion
  - [x] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [x] tbl-results-summary
  - [ ] tbl-predictors
  - [x] palaeodist figs
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
  - [ ] Replace "see below" etc. with section links
- [x] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

-->

@@

The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@cite].
These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@cite] and eventually pre-domestication cultivation [@cite].
As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.

@@

Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
In the first sense, it has been used most extensively to model the range of humans and other hominin species [e.g. @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—or more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
These applications can be distinguihed from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
In an early application to West Asia, @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic sites on the Konya Plain [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [e.g. @CremaEtAl], though not as yet applied to West Asia.

@@

The alternative approach is to train the model using contemporary occurrence and environmental data and then use palaeoenvironmental data to 'hindcast' its predictions backwards in time.
Like @FranklinEtAl2015, we view the hindcasting approach as more promising, because training datasets for both occurrences and environment are far more readily available, complete and reliable for the present than the past.
There is some scepticism in the ecological niche modelling literature about the ability of such models to make accurate predictions in unknown environments [like the past, @FranklinEtAl2015], 
but here the hindcasting approach also presents an opportunity:
it reserves archaeological occurrence data as an independent dataset that can be used to assess the retrodictive performance of the model.
This possibily was suggested by @FranklinEtAl2015 but to our knowledge our study represents the first attempt to actually do so.

@@

```{r data-flora}
# Start with flora present at archaeological sites
if (!file_exists(derived_data("flora.Rdata"))) {
  flora <- read_tsv(
    raw_data("swasia_neolithic_flora.tsv"),
    col_types = cols(
      source = col_character(),
      source_id = col_character(),
      source_site_name = col_character(),
      site_name = col_character(),
      latitude = col_double(),
      longitude = col_double(),
      phase = col_character(),
      phase_description = col_character(),
      phase_code = col_character(),
      age_start = col_integer(),
      age_end = col_integer(),
      taxon_source = col_character(),
      n = col_double(),
      prop = col_double(),
      reference = col_character(),
      taxon_detail = col_character(),
      taxon = col_character(),
      genus = col_character(),
      family = col_character(),
      category = col_character(),
      founder_crop = col_character(),
      edibility = col_character(),
      grass_type = col_character(),
      legume_type = col_character()
    )
  ) |>
    rename(taxon_group = "taxon", taxon = "taxon_detail") |>
    # Assign to archaeological periods
    mutate(
      age_mid = age_end + ((age_start - age_end) / 2),
      period = cut(
        -age_mid, 
        breaks = -c(archaeo_periods$start_bp, 0), 
        labels = archaeo_periods$period,
        ordered_result = TRUE
      ),
      climate_period = cut(
        -age_mid, 
        breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
        labels = climate_periods[climate_periods$code != "cur",]$code,
        ordered_result = TRUE
      )
    ) |>
    # Filter to Epipal/Neo
    filter(period %in% epipal | period %in% ppn) |>
    # Group by taxon
    drop_na(taxon) |>
    nest(.by = c(taxon), .key = "archaeo") |>
    # Normalise species names and match to the GBIF Backbone Taxonomy
    mutate(
      # Spot corrections
      taxon = case_match(
        taxon,
        "Vicia narbonense" ~ "Vicia narbonensis", # N=3
        "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
        "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
        "Lens orientalis" ~ "Vicia orientalis", # N=2
        "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
        "Stipa capensis" ~ "Stipa dregeana", # N=2
        .default = taxon
      ),
      taxon_original = taxon,
      taxon = map_chr(taxon, gbif_accepted_name),
      .before = archaeo
    ) |>
    # Drop unmatched (mostly not identified to species)
    drop_na(taxon) |>
    # Recode domesticates as their wild progenitors.
    mutate(
      taxon = case_match(
        taxon,
        "Cicer arietinum" ~ "Cicer reticulatum",
        "Hordeum vulgare" ~ "Hordeum spontaneum",
        "Vicia lens" ~ "Vicia orientalis",
        "Linum usitatissimum" ~ "Linum bienne",
        "Triticum aestivum" ~ "Triticum turgidum dicoccum",
        "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
        "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
        "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
        .default = taxon
      )
    ) |>
    # Regroup by canonical taxon
    summarise(
      taxon_aliases = list(taxon_original[taxon_original != taxon]),
      archaeo = list(bind_rows(archaeo)),
      .by = taxon
    ) |>
    mutate(
      n_archaeo = map_int(archaeo, nrow)
    ) |>
    filter(n_archaeo > min_n_archaeo) |>
    # Add wild progenitors of wheats
    bind_rows(
      tibble(
        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"),
        taxon_original = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
      )
    )

  saveRDS(flora, derived_data("flora.Rdata"))
}
flora <- readRDS(derived_data("flora.Rdata"))
```

```{r data-archaeo-sites}
archaeo_sites <- flora |>
  unnest(archaeo) |>
  summarise(
    latitude = first(latitude), 
    longitude = first(longitude), 
    .by = c(site_name, period)
  ) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = latlong)
```

@@

The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@COMPAG].
We did not attempt to distinguish between the source of the remains [cf. @WallaceEtAl2018];
archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
However, we assume that there presence at a site of human settlement at least implies that they were part of the wider ecosystem that supported habitation there.

Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-results-summary).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.

@@

<!-- TODO generate GBIF occurrence citation:
https://docs.ropensci.org/rgbif/articles/getting_occurrence_data.html -->

```{r data-n-occ} 
if (!file_exists(derived_data("flora_n_occ.Rdata"))) {
  message("Reacquiring GBIF occurrence counts...")
  flora <- flora |>
    mutate(
      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                            hasCoordinate = TRUE, 
                                            hasGeospatialIssue = FALSE,
                                            geometry = bbox_wkt(w_asia)))
    )
  saveRDS(flora, derived_data("flora_n_occ.Rdata"))
}
flora <- readRDS(derived_data("flora_n_occ.Rdata"))

@@

```{r data-occ}
# Cleaning pipeline based on:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
if (!file_exists(derived_data("flora_occ.Rdata"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, filter, occurrenceStatus  == "PRESENT"),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 1000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    }),
    # Update occ count after cleaning
    n_occ = map_int(occ, nrow)
  )

@@

  saveRDS(flora, derived_data("flora_occ.Rdata"))
}

flora <- readRDS(derived_data("flora_occ.Rdata"))
```

Georeferenced occurrence data for West Asia was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
GBIF was cleaned to removed fossil occurrences, recorded absences, and records with missing, imprecise (>1 km uncertainty), or invalid coordinates.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what accounted for such a disproportionately high number of records.

@@

## Predictor data {#sec-predictors}

We modelled the occurrence of species as a function of 24 geospatial predictor variables.
These included:

@@

For hindcasting, we used reconstructed bioclimatic data for three key climatologies generated from downscaled paleoclimate simulations from the HadCM3 general circulation model [@FordhamEtAl2017; @BrownEtAl2018]:
the Bølling–Allerød (c. 14.7–12.9 ka), the Younger Dryas (c. 12.9–11.7 ka), and the Early Holocene (11.7–8.3 ka).
Terrain and soil predictors were held constant, since reconstructions of these variables in the past are not available at sufficient scale.
It is unlikely that either macroscale topography or soil characteristics have altered significantly over the period of time considered here, so we assume that this does not degrade model performance, and may in fact benefit it by providing 'anchoring' predictors that are independent of climate change.

For training, test predictions, and archaeological predictions predictor data was left in its native projection and resolution.
For hindcast palaeodistributions, it was transformed to common equal-area projection and resolution of `r pred_res / 1000` km.

@@

```{r niche-data}
# Split training, test (for model assessment) and cross-validation sets (for
# tuning)
flora <- flora |>
  mutate(
    # De-sf
    occ = map(occ, \(x) select(tibble(x), -geometry)),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
  )

@@

For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our 24 predictor variables (@sec-predictors).
We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (¾) and test (¼) partitions.

@@

We trained Random Forest models for `r nrow(flora)` taxa using contemporary occurrence data from GBIF, a random sample of background points, and the predictor variables described in @sec-predictors.
Substituting the "current" climate predictors for those derived from palaeoclimatic simulations [@BrownEtAl2018], we could then generate hindcast predictions for reconstructed past environments in `r english(nrow(climate_periods))` key climate periods – a total of `r nrow(flora) * nrow(climate_periods)` modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the appendix and accompanying material.

@@

```{r fig.height=10, fig-palaeodist-barely-levant, cache=TRUE}
#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
# TODO: why so slow?
palaeodist_hordeum <- do.call(c, c(
  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
))
